\myfont Implementation and evaluation of register allocation for translating LLVM-{}- to x86-64

### 2024-01-12 article.tex
Cite variadic arguments

To generally map each LLVM-{}- instruction to what amounts to the 'core' function in x86, consider Table \ref{tab:trans}.

\begin{table}[H]
  \centering
\begin{tabular}{|>{\tt}l|>{\tt}l|l|} 
 \hline
 LLVM-{}- & x86 & Registers affected\\
 \hline
add      & addx \%src, \%dst & \{\}\\
alloca   & subq\ \ \_\_\_, \%rsp & \{\texttt{\%rsp}\}\\
allocan  & subq\ \ \_\_\_, \%dst & \{\texttt{\%rsp}\}\\
and      & andx \%src, \%dst & \{\}\\
ashr     & sarx \%src, \%dst & \{\}\\
bitcast  & movx \%src, \%dst & \{\}\\
call     & \textrm{\textit{see \ref{impl-call}}} & \{\texttt{\%rax}, \texttt{\%rcx}, \texttt{\%rdx}, \texttt{\%rsi}, \texttt{\%r8-\%r11}\}\\
gep      & \textrm{\textit{see \ref{impl-gep}}} & \{\}\\
icmp     & cmpx \%src, \%dst & \{\texttt{\%rFLAGS}\}\\
load     & movx (\%src), \%dst & \{\}\\
lshr     & shrx \%src, \%dst & \{\}\\
mul      & imulx \%src, \%dst & \{\texttt{\%rax}\}\\
or       & orx \%src, \%dst & \{\}\\
phi      & \textrm{\textit{see \ref{impl-phi}}} & \{\}\\
ptrtoint & movx \%src, \%dst & \{\}\\
sdiv     & idivx \%src, \%dst & \{\texttt{\%rax}, \texttt{\%rdx}\}\\
select   &  & \{\}\\
sext     & movx \%src, \%dst & \{\}\\
shl      & shl \%src, \%dst & \{\}\\
srem     & idivx \%src, \%dst & \{\texttt{\%rax}, \texttt{\%rdx}\}\\
store    & movx \%src, (\%dst) & \{\}\\
sub      & subx \%src, \%dst & \{\}\\
trunc    & movx \%src, \%dst & \{\}\\
udiv     & divx \%src, \%dst & \{\texttt{\%rax}, \texttt{\%rdx}\}\\
urem     & divx \%src, \%dst & \{\texttt{\%rax}, \texttt{\%rdx}\}\\
xor      & xorx \%src, \%dst & \{\}\\
zext     & movx \%src, \%dst & \{\}\\
 \hline
\end{tabular}
\caption{LLVM-{}- to x86 translation table}\label{tab:trans}
\end{table}

 %add & addx & \{\}\\
 %\hline
 %mul & imul & \{ \%rax \} \\
 %\hline
 %sdiv & idivx & \{\%rax, \%rdx\} \\
 %\hline
 %srem & idivx & \{\%rax, \%rdx \\
 %\hline
   %call & callx & \{\%rax, \textit{caller saved registers}\} \\

@@

The \mintinline{llvm}{call} operation is by far the most complicated one. Mostly because of calling conventions,
The \texttt{\%rax} register is zeroed before

\subsubsection{Implementing \mintinline{llvm}{gep} instructions}\label{impl-gep}
GEP instructions are used to perform address calculations for accessing elements in  structures or arrays. They are particularly useful for efficiently addressing elements in complex data structures. For example, code accessing fields like line 8 in Figure \ref{fig:vector.c}, would  translate to the snippet seen in Figure \ref{fig:vector.ll}.

>>>>

To map each LLVM-{}- instruction to what amounts to the 'core' function in x86, consider Table \ref{tab:trans}. Instructions for which the size varies, the suffix is denoted with \texttt{x} instead. This is then replaced with the appropriate size suffix depending on the size denoted in the LLVM-{}- instruction. For example, most binary operations move the left-hand operand into \texttt{\%rax} and the right-hand operand into  \texttt{\%rcx}, after which the associated instruction from the table is performed. For x86, most operations are done in place, so for the most part \texttt{\%rax} is modified and moved to the destination after.

\begin{table}[H]
  \centering
\begin{tabular}{|>{\tt}l|>{\tt}l|l|} 
 \hline
 LLVM-{}- & x86 & Registers affected\\
 \hline
add      & addx \%src, \%dst & \{\}\\
alloca   & subq \$n, \%rsp & \{\texttt{\%rsp}\}\\
allocan  & subq \$n, \%rsp & \{\texttt{\%rsp}\}\\
and      & andx \%src, \%dst & \{\}\\
ashr     & sarx \%src, \%dst & \{\}\\
bitcast  & movx \%src, \%dst & \{\}\\
call     & \textrm{\textit{see \ref{impl-call}}} & \{\texttt{\%rax}, \texttt{\%rcx}, \texttt{\%rdx}, \texttt{\%rsi}, \texttt{\%r8-\%r11}\}\\
gep      & \textrm{\textit{see \ref{impl-gep}}} & \{\}\\
icmp     & cmpx \%src, \%dst & \{\texttt{\%rFLAGS}\}\\
load     & movx (\%src), \%dst & \{\}\\
lshr     & shrx \%src, \%dst & \{\}\\
mul      & imulx \%src, \%dst & \{\texttt{\%rax}\}\\
or       & orx \%src, \%dst & \{\}\\
phi      & \textrm{\textit{see \ref{impl-phi}}} & \{\}\\
ptrtoint & movx \%src, \%dst & \{\}\\
sdiv     & idivx \%src, \%dst & \{\texttt{\%rax}, \texttt{\%rdx}\}\\
select   &  & \{\}\\
sext     & movx \%src, \%dst & \{\}\\
shl      & shl \%src, \%dst & \{\}\\
srem     & idivx \%src, \%dst & \{\texttt{\%rax}, \texttt{\%rdx}\}\\
store    & movx \%src, (\%dst) & \{\}\\
sub      & subx \%src, \%dst & \{\}\\
trunc    & movx \%src, \%dst & \{\}\\
udiv     & divx \%src, \%dst & \{\texttt{\%rax}, \texttt{\%rdx}\}\\
urem     & divx \%src, \%dst & \{\texttt{\%rax}, \texttt{\%rdx}\}\\
xor      & xorx \%src, \%dst & \{\}\\
zext     & movx \%src, \%dst & \{\}\\
 \hline
\end{tabular}
\caption{LLVM-{}- instruction to x86 translation table}\label{tab:trans}
\end{table}

\begin{table}[H]
  \centering
\begin{tabular}{|>{\tt}l|>{\tt}l|l|} 
 \hline
 LLVM-{}- & x86 & Registers affected\\
 \hline
add      & addx \%src, \%dst & \{\}\\
alloca   & subq \$n, \%rsp & \{\texttt{\%rsp}\}\\
allocan  & subq \$n, \%rsp & \{\texttt{\%rsp}\}\\
and      & andx \%src, \%dst & \{\}\\
ashr     & sarx \%src, \%dst & \{\}\\
bitcast  & movx \%src, \%dst & \{\}\\
call     & \textrm{\textit{see \ref{impl-call}}} & \{\texttt{\%rax}, \texttt{\%rcx}, \texttt{\%rdx}, \texttt{\%rsi}, \texttt{\%r8-\%r11}\}\\
gep      & \textrm{\textit{see \ref{impl-gep}}} & \{\}\\
icmp     & cmpx \%src, \%dst & \{\texttt{\%rFLAGS}\}\\
load     & movx (\%src), \%dst & \{\}\\
lshr     & shrx \%src, \%dst & \{\}\\
mul      & imulx \%src, \%dst & \{\texttt{\%rax}\}\\
or       & orx \%src, \%dst & \{\}\\
phi      & \textrm{\textit{see \ref{impl-phi}}} & \{\}\\
ptrtoint & movx \%src, \%dst & \{\}\\
sdiv     & idivx \%src, \%dst & \{\texttt{\%rax}, \texttt{\%rdx}\}\\
select   &  & \{\}\\
sext     & movx \%src, \%dst & \{\}\\
shl      & shl \%src, \%dst & \{\}\\
srem     & idivx \%src, \%dst & \{\texttt{\%rax}, \texttt{\%rdx}\}\\
store    & movx \%src, (\%dst) & \{\}\\
sub      & subx \%src, \%dst & \{\}\\
trunc    & movx \%src, \%dst & \{\}\\
udiv     & divx \%src, \%dst & \{\texttt{\%rax}, \texttt{\%rdx}\}\\
urem     & divx \%src, \%dst & \{\texttt{\%rax}, \texttt{\%rdx}\}\\
xor      & xorx \%src, \%dst & \{\}\\
zext     & movx \%src, \%dst & \{\}\\
 \hline
\end{tabular}
\caption{LLVM-{}- instruction to x86 translation table}\label{tab:trans}
\end{table}

@@

The \mintinline{llvm}{call} operation is by far the most complicated one. Mostly because of calling conventions,
The \texttt{\%rax} register is zeroed before every call as its lowermost byte register \texttt{\%al} is used to indicate the number of variadic arguments \cite[25]{sysv}.

\subsubsection{Implementing \mintinline{llvm}{getelementptr} instructions}\label{impl-gep}
GEP instructions are used to perform address calculations for accessing elements in  structures or arrays. They are particularly useful for efficiently addressing elements in complex data structures. For example, code accessing fields like line 8 in Figure \ref{fig:vector.c}, would  translate to the snippet seen in Figure \ref{fig:vector.ll}.
